Score-based tests of measurement invariance: Use in practice

### 2013-10-29 frontiers/frontiers.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@290 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
small improvements
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@291 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

This section contains background and discussion of the  proposed statistics as
applied to SEM; for a more detailed account, see \citeA{MerZei13} and
\citeA{MerFanZei}.  For details on the statistics' application to general
statistical models, see \citeA{zeihor07}.

@@

In this section, we demonstrate how the above tests can be carried out in~R, 
using the package \pkg{lavaan} \cite{lavaan11} for model estimation and \pkg{strucchange} 
\cite{ZeiLei02,Zei06} for testing.  We use data from \citeA{FroFan11} concerning 
the applicability of adult gratitude scales to youth, available in the~\proglang{R} 
package \emph{psychotools} \cite{zeilei12}.  The data consist of responses to three adult 
gratitude scales from $n=1401$ youth aged 10--19 years. The original authors 
were specifically interested in whether the scales were measurement invariant 
across age.  Because the sample size at each age was unbalanced, the authors 
created age groups of approximately equal sample size.  In the examples below, 
we test for measurement invariance across these age groups.

@@

If we are interested in measurement invariance violations that are 
monotonic with the age groups, it is perhaps simplest to treat the 
age groups as continuous.  In doing so, we can use the statistics 
from Equations~\eqref{eq:dmax}, ~\eqref{eq:cvm}, and~\eqref{eq:maxlm}.  
That is, we can fit a model whose parameters 
are restricted to be equal across all individuals and then examine 
how individuals' scores $s(\hat{\theta};x_{i})$ fluctuate with their
age (where age ties are broken arbitrarily, using the original order
of the observations within each age group).
This is demonstrated below, with similar code being useful when one is 
testing for measurement invariance w.r.t.\ truly continuous variables.

@@

Finally, if the user anticipates multiple calculations of the
$\max\mathit{LM}_o$ statistic for a specific dataset, it is possible
to save time by simulating critical values once and re-using them for
multiple tests.  We can use the \code{ordL2BB()} command to generate
critical values and store them in an object \code{mLMo}, say.
Then, this object can be employed to obtain the test statistic in the usual manner.
<<ordsave, eval=FALSE>>=
mLMo <- ordL2BB(yg$agegroup)
maxlmo <- sctest(restr, order.by = yg$agegroup, parm = 1:4,
  vcov = "info", functional = mLMo)
@ 
The \code{ordL2BB()} command automatically generates critical values
for testing 1 to 20 parameters at a time. If only a smaller number
of parameters (e.g., only up to 6) is to be tested, some computation time can be
saved by setting the \code{nproc} argument accordingly (e.g., \code{nproc = 1:6}).
In the same way, \code{nproc} can be employed to simulate higher-dimensional
fluctuation processes suitable for testing more parameters. One can re-use
\code{mLMo} in this manner for further tests of the youth gratitude data.
Critical values must be resimulated for new data, however, because they depend on the
proportion of individuals observed at each level of the ordinal
variable (denoted $t_\ell$ for Equation~\eqref{eq:maxlmo}).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Power and Type I error were examined across three sample sizes
($n=120,480,960$), three numbers of categories ($m=4,8,12$), and
17 magnitudes of invariance violations (described in the following sentences). 
The measurement invariance violations began at level $1+m/2$ of the
auxiliary variable $V$ and were consistent thereafter: individuals below
level $1+m/2$ of $V$ deviated from individuals at or above level $1+m/2$ 
by $d$ times the parameters' asymptotic standard errors
(scaled by $\sqrt n$), with $d=0,0.25,0.5,...,4$. 
For each combination of sample size ($n$) $\times$ violation
magnitude ($d$) $\times$ violating parameter $\times$ categories ($m$), 5000 datasets were
generated and tested.  Statistics from Equations~\eqref{eq:wdm}, ~\eqref{eq:maxlmo}
and~\eqref{eq:lmuo} were examined. As mentioned previously,
~\eqref{eq:lmuo} is asymptotically equivalent to the usual likelihood
ratio test.  Thus, this statistic provides information about the
relative performance of the ordinal statistics vs.\ the LRT.

@@

In the reminder of the paper, we first compare
these tests to traditional tests of measurement invariance.  We then 
discuss test extension to other fit functions and to other specialized
models.

>>>>

This section contains background and discussion of the  proposed statistics as
applied to structural equation models (SEMs); for a more detailed account, see \citeA{MerZei13} and
\citeA{MerFanZei}.  For details on the statistics' application to general
statistical models, see \citeA{zeihor07}.

@@

In this section, we demonstrate how the above tests can be carried out in \proglang{R}, 
using the package \pkg{lavaan} \cite{lavaan11} for model estimation and \pkg{strucchange} 
\cite{ZeiLei02,Zei06} for testing.  We use data from \citeA{FroFan11} concerning 
the applicability of adult gratitude scales to youth, available in the~\proglang{R} 
package \pkg{psychotools} \cite{zeilei12}.  The data consist of responses to three adult 
gratitude scales from $n=1401$ youth aged 10--19 years. The original authors 
were specifically interested in whether the scales were measurement invariant 
across age.  Because the sample size at each age was unbalanced, the authors 
created age groups of approximately equal sample size.  In the examples below, 
we test for measurement invariance across these age groups.

@@

If we are interested in measurement invariance violations that are 
monotonic with the age groups, it is perhaps simplest to treat the 
age groups as continuous.  In doing so, we can use the statistics 
from Equations~\eqref{eq:dmax}, \eqref{eq:cvm}, and~\eqref{eq:maxlm}.  
That is, we can fit a model whose parameters 
are restricted to be equal across all individuals and then examine 
how individuals' scores $s(\hat{\theta};x_{i})$ fluctuate with their
age (where age ties are broken arbitrarily, using the original order
of the observations within each age group).
This is demonstrated below, with similar code being useful when one is 
testing for measurement invariance w.r.t.\ truly continuous variables.

@@

Finally, if the user anticipates multiple calculations of the
$\max\mathit{LM}_o$ statistic for a specific dataset, it is possible
to save time by simulating critical values once and re-using them for
multiple tests.  We can use the \code{ordL2BB()} function to generate
critical values and store them in an object \code{mLMo}, say.
Then, this object can be employed to obtain the test statistic in the usual manner.
<<ordsave, eval=FALSE>>=
mLMo <- ordL2BB(yg$agegroup)
maxlmo <- sctest(restr, order.by = yg$agegroup, parm = 1:4,
  vcov = "info", functional = mLMo)
@ 
The \code{ordL2BB()} command automatically generates critical values
for testing 1 to 20 parameters at a time. If only a smaller number
of parameters (e.g., only up to 6) is to be tested, some computation time can be
saved by setting the \code{nproc} argument accordingly (e.g., \code{nproc = 1:6}).
In the same way, \code{nproc} can be employed to simulate higher-dimensional
fluctuation processes suitable for testing more parameters. One can re-use
\code{mLMo} in this manner for further tests of the youth gratitude data.
Critical values must be resimulated for new data, however, because they depend on the
proportion of individuals observed at each level of the ordinal
variable (denoted $t_\ell$ for Equation~\eqref{eq:maxlmo}).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Power and Type I error were examined across three sample sizes
($n=120,480,960$), three numbers of categories ($m=4,8,12$), and
17 magnitudes of invariance violations (described in the following sentences). 
The measurement invariance violations began at level $1+m/2$ of the
auxiliary variable $V$ and were consistent thereafter: individuals below
level $1+m/2$ of $V$ deviated from individuals at or above level $1+m/2$ 
by $d$ times the parameters' asymptotic standard errors
(scaled by $\sqrt n$), with $d=0,0.25,0.5,...,4$. 
For each combination of sample size ($n$) $\times$ violation
magnitude ($d$) $\times$ violating parameter $\times$ categories ($m$), 5000 datasets were
generated and tested.  Statistics from Equations~\eqref{eq:wdm}, \eqref{eq:maxlmo}
and~\eqref{eq:lmuo} were examined. As mentioned previously,
\eqref{eq:lmuo} is asymptotically equivalent to the usual likelihood
ratio test.  Thus, this statistic provides information about the
relative performance of the ordinal statistics vs.\ the LRT.

@@

In the remainder of the paper, we first compare
these tests to traditional tests of measurement invariance.  We then 
discuss test extension to other fit functions and to other specialized
models.
